Convergence Studies of Mass Transport in Disks with Gravitational Instabilities. \\ I. The Constant Cooling Time Case

### 2011-06-08 Paper_1.tex
Tom's submitted version w/ pared down abstract.
	modified:   Paper_1.tex
	modified:   figures/Figure1.eps
	modified:   figures/Figure3.eps
	modified:   figures/Figure5.eps
	modified:   figures/Figure7.eps
	modified:   general.bib

\usepackage{natbib}
\usepackage{color}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{grffile}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{amssymb}

@@

\title{Convergence Studies on Mass Transport in Disks with Gravitational Instabilities.
 \\
I. The Constant Cooling Case}
\author{Scott Michael}
\affil{Astronomy Department, Indiana University, Bloomington, IN 47405}
\email{scamicha@astro.indiana.edu}

@@

\author{Aaron Boley}
\affil{Department of Astronomy, University of Florida, Gainesville, FL 32611}
\email{aaron.boley@gmail.com}

@@

We conducted a convergence study of the azimuthal resolution in 3D hydrodynamic simulations of a protostellar disk susceptible to gravitational instabilities (GIs). 
The disk was primarily studied during its asymptotic state, when
heating and cooling are in balance. The cooling was treated as a constant global cooling time.
Four simulations were conducted, identical except for the number of azimuthal grid 
points, $l_{max} = $ 64, 128, 256 and 512, used in the calculations. 
The strengths of the azimuthal modes, gravitational torques, and the effective \citet{shakura1973} $\alpha$ 
arising from gravitational stresses were determined for each resolution.  We examined the resolution-dependence of each quantity
and evaluated both the converged modal amplitude distribution and the effective value of $\alpha$
as $l_{max} \rightarrow \infty$.   Non-axisymmetric amplitude was found in all modes and 
the amplitude summed over all modes was essentially independent of resolution.
The modal distribution of amplitudes, however, was significantly affected by resolution.  
The size of the azimuthal grid, $l_{max}$,
limits the number of modes that can be accurately measured to $l_{max}/2$.  Higher resolution simulations therefore
have more modes available to them.  Non-axisymmetric structure is present in all available modes.  As a consequence,
higher resolution simulations have a larger fraction of their total amplitude in high-order modes.
High-order modes
are more confined to act on local structure than low-order modes, therefore as the resolution increases and low-order amplitudes decrease the total gravitational stresses decreases as well.
As a result, higher resolution simulations experience 
weaker gravitational torques than lower resolution simulations.   The effective  $\alpha$
also depends upon the magnitude of the stresses.  Thus
$\alpha_{\rm{eff}}$ is also a function of grid resolution, decreasing as  $l_{max}$ increases.
The converged $\alpha_{\rm{eff}}$ is consistent with
predictions from an analytic local theory for thin disks by \citet{gammie2001}, 
but only over many dynamic times when averaged over a substantial volume of the disk.
Results obtained in the $l_{max} = 512$ simulations differ with converged values
by less than 20\%.

@@

Detailed three-dimensional hydrodynamic modeling of protostellar disks has provided considerable insight into the physical and thermal states of these disks.  However,  computing resources limit these studies to snapshots spanning timescales much shorter than a disk's evolutionary lifetime. The level of spatial resolution (and hence temporal resolution, due to the Courant condition) used in calculations can play a deterministic role in interpretations of models; while lower resolutions allow more extensive spatial and temporal studies, this benefit can come at the sacrifice of important physics.  Convergence studies provide a means 
of deriving important outcomes without the expense of very high resolution simulations.

In contrast to sophisticated calculations that include detailed physics, 
approaches exist that replace detailed physics with statistical parameters that encapsulate this physics.  This
allows for calculations covering much longer timeframes.  Both detailed and 
parametric
approaches can play important roles in broadening our understanding of complex nonlinear systems, provided a connection between the detailed physics and statistical parameters can be demonstrated and understood. In what follows, we perform convergence 
tests on hydrodynamical disk simulations to examine the applicability of an $\alpha$-disk 
formulation \citep{shakura1973} to  the
evolution of protostellar disks subject to gravitational instabilities.

%Here $\alpha$ is a dimensionless parameter that
%provides a measure of viscosity, without providing a detailed specification of the viscous mechanism.

Gravitational instabilities (GIs) can play an important, and at times dominant, role in driving the structural 
evolution of  protostellar disks \citep[for reviews, see][]{durisen2005, durisen2010}. {\bf (We need the Durisen 2010 (2011?)
review article reference)}
Thermal processes play the primary role in regulating the amplitude and outcome of these instabilities \citep{pickett1998,pickett2000,mejia2005,nelson1998,nelson2000a}. A disk's susceptibility to GIs can be parameterized by the Toomre $Q$-parameter \citep{toomre1981};  $Q = c_s \kappa / \pi G \Sigma$, where $c_s$ is the sound speed, $\kappa$ is the epicyclic frequency ($\sim$ the rotation frequency 
$\Omega$ in a nearly Keplerian disk), and $\Sigma$ is the disk surface mass density. For $Q \al$  1.5 to 1.7,  small density perturbations in a disk grow exponentially on a time scale comparable to the rotation period. These perturbations manifest themselves as multi-arm spirals with a predominantly trailing pattern that transports angular momentum outward by gravitational torques \citep{larson1984,boss1984,durisen1986,papaloizou1991, laughlin1994,nelson1998,pickett1998}.

@@

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime \citep{nelson1998, pickett1998, lodato2004,boley2006,cossins2009}.  Even if the initial growth is dominated by one mode, numerous modes usually appear in the nonlinear regime. As a consequence, if a disk achieves a quasi-steady balance between heating and cooling, referred to as the {\it asymptotic state},
then power will exist at all $m$-values resolved by a given numerical method \citep{mejia2005,boley2006}.  
In the asymptotic state, GI-active disks develop overlapping density structures of different strengths, geometries and coherence, producing strongly nonlinear turbulence that pervades the entire disk \citep[e.g.][]{pickett2003, mejia2005,boley2006}.
\citet{gammie2001} refers to this state as {\it gravitoturbulence}. Gravitoturbulence provides a possible bridge between the detailed physics of GIs and the viscous transport parameter $\alpha$.  Here convergence studies are important because 
higher order modes dissipate energy on shorter wavelengths than lower order modes.  The mix of low and high order modes available to a disk is set by the resolution of the simulation. {\bf Multiple issues with this paragraph. Didn't want to rewrite unilaterally. 1) In the first couple sentences the description of mode growth needs some tweaking, many modes appear in the linear regime, and although we have seen amplitude at all m values I don't know that it's a given. 2) I wouldn't describe the structures formed by GI activity as ``strongly nonlinear turbulence''. 3) In reviewing Gammie 2001 it seems that he is actually referring to a turbulent cascade (a fact I hadn't fully appreciated until now) I think it's pretty clear that the effective alpha we see is, in fact, not due to a turbulent cascade. i.e. alpha is driven by low order coherent modes. We should discuss this at some point SM}

\citet{shakura1973} proposed turbulence as the primary source of effective viscosity in gas disks
and parameterized turbulent transport in a steady-state disk with the  parameter $\alpha$.  The source of the 
turbulence in their $\alpha$-disk formulation is unspecified. Although their original work was directed towards X-ray bright
accretion disks around black holes, this heuristic approach has proven fruitful in many astrophysical situations by allowing for analytic solutions and relatively easy numerical modeling over time-spans inaccessible to detailed modeling. In the 
$\alpha$-disk formalism, when the accretion rate $\dot{M}$ is constant in both time and radius, $\dot{M} = 3\pi\alpha c_s \Sigma H$ where $\Sigma$ is the surface density and $H$ represents
the disk scale height. The $\alpha$ formalism makes no assumption about the nature of the angular momentum transport, only that {\it angular momentum transport
and heating are dominated by local processes}. \citet{lin1987,lin1990} were among the first to suggest
that GI-induced transport could be described within a viscous
framework.  However, the  validity of an $\alpha$-disk picture requires self-consistency of local energy dissipation. 
The locality of dissipation in GI-active disks has not been rigorously tested and remains an open question.

Several studies support the idea that angular momentum transport by GIs in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006}.  
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in balance, i.e., disks subject to gravitoturbulence,
 it may be valid to represent GI-induced transport by an {\it effective viscosity}, a local process \citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.  In particular, 
an $\alpha$-viscosity may be consistent
with transport in GI active disks for disk to star mass ratio $M_d/M_* \le 0.2$ -- 0.3; GI-active disks
in systems with larger $M_d/M_*$ are dominated by lower-order modes that act in a more global sense and thus cannot be well represented by an $\alpha$-approach
\citep{lodato2004, cossins2009, vorobyov2010}.

\citet{gammie2001} derived an effective alpha for the case of a thin 2D gravitationally unstable disk where the heating caused by GIs was balanced by the local cooling rate.  
Assuming a razor thin gravitationally unstable disk where the heating caused by
GIs was balanced by a local cooling rate of $t_{cool}\Omega$.
Gammie found
\begin{equation}
\alpha_{\rm{eff}} = [\gamma(\gamma - 1) \slantfrac{9}{4}\Omega t_{cool}]^{-1},
\label{eq:gammie}
\end{equation}
where $\gamma$ is the two dimensional adiabatic index and $t_{cool}$ is a cooling parameter. 
The two dimensional index relates
to the three dimensional index, $\Gamma$, by 
$\gamma = (3\Gamma - 1) / (\Gamma + 1)$
for a non-self-gravitating
disk and
$\gamma = 3 - 2/\Gamma $
for a strongly self gravitating disk.   \cite{lodato2004} derived a similar result when considering three dimensional protoplanetary disks.
Motivated by
the results from \cite{laughlin1996}, \cite{gammie2001}, and \cite{lodato2004, lodato2005}, which demonstrated differences in the manifestation of high and low-order, \citet{kratter2008} 
proposed $\alpha_{\rm{eff}} = (\alpha_{short}^2 + \alpha_{long}^2)^{1/2}$. Here $\alpha_{short}$ and $\alpha_{long}$
are functions of $Q$ and $M_d/\left( M_d+M_*\right)$ representing 
the different wavelength regimes expected to dominate for different values of the mass ratio and cooling rate. 
Fully 3D simulations \citep[e.g.][]{lodato2004, cossins2009} suggest that, at least in the case of imposed cooling, $\alpha_{\rm{\rm{eff}}}$ roughly converges to the value predicted by \citeauthor{gammie2001}.

We address these, and similar questions, using a grid-based, finite-difference 3D hydrodynamics code
with self gravity.
For computational grids of a fixed {\it physical} size,
grids with more elements provide finer meshes, and thus a more accurate solution. 
However, increased accuracy is offset by computational penalties.
The number of equations solved at each time step 
increases as the grid size is increased.  At the same time, the Courant condition 
forces shorter time steps.  A mesh convergence study 
provides a common approach to balancing accuracy and computing resources.
In the following sections we follow the evolution of a disk subject to GIs
using four different
resolutions in the azimuthal direction.  The number of azimuthal grid points used in simulations
may be particularly important, because higher azimuthal resolution 
allows non-axisymmetric amplitude to spread to higher-order modes 
that behave more locally. We test this by examining the amplitude in various 
modes to see how they are affected by the choice of grid.
In the process, 
we report the results of a convergence study designed to determine whether GI-active disks, chaotic by nature, achieve statistical equilibrium If this is the case, we investigate how closely these disks approximate the gravitoturbulence of \citet{gammie2001} and examine the distribution of $\alpha_{\rm{eff}}$. To this end, we follow the evolution of a GI-active disk, subject to constant cooling, in its asymptotic state, where cooling and heating are essentially balanced.  Simulations are performed using four different resolutions in the azimuthal direction.

The balance of this paper is organized as follows.  Section 2 provides the details of the numerical approach and defines the models.  An analysis of the modal structure of the simulations 
is found in \S3.1, the gravitational torques arising from these modes follows in \S3.2.  Effective converged alphas are 
discussed in \S3.3.  Section 3.4 deals with the degree to which selected grid sizes lead to results consistent with
convergence.  Fragmentation is discussed in \S3.5.  A summary of results is found in \S4.

%We cannot yet follow the entire dynamical evolution of a 
%protoplanetary disk nor
%model all the radiative and particle processes of interest at once in full 3D. However, 
%by following GI simulations to their asymptotic states, when these exist, we
%can extrapolate the consequences over much longer times (e.g., Boley et al 2006).

@@

Simulations were conducted with the three-dimensional hydrodynamics code used by the Indiana University Hydrodynamics Group in several previous studies \citep{pickett1998, pickett2000, mejiaphd2004, mejia2005, boley2006, cai2008}.  This code uses a second-order, explicit, Eulerian scheme to solve Poisson's equation, an equation of state, and the equations of hydrodynamics in conservative form on a uniform cylindrical grid.  The code assumes mirror symmetry about the equatorial plane. It includes both self-gravity and artificial viscosity; the latter serves to 
mediate shocks. Source and flux terms \citep{norman1986} are computed separately in an explicit, second-order time integration \citep{albada1982,christodoulou1991,yangphd1992}.

We assume an equation of state $P = (\gamma -1)\epsilon$, where $P$ is the pressure, $\epsilon$ is the internal energy density, and $\gamma$ is the ratio of specific heats, given here by $\gamma = 5/3$.  The model disks are cooled by decreasing their internal energy according to the prescription $d\epsilon/dt = \epsilon / t_{cool}$, where $t_{cool}$ is the global cooling time in Outer Rotation Periods (ORP).  For the grid resolutions used here, one ORP is defined as the rotational period at radial zone 200.

%\subsection{Models}

Because this is a polytropic disk with an idealized cooling time, only the ratio of disk to stellar mass, given here by 0.153, is specified.  In what follows, quantities specified with real physical units are derived assuming a 1 $M_\odot$ central star and a 0.153  $M_\odot$ disk. In this case, the ORP is defined at $r = 33$ AU and one ORP $\approx$ 303 years.  Models have $t_{cool} = 2$ ORP. The disk has an initial surface density $\Sigma \sim r^{ -1}$, and initial inner and outer radii of 2.3 and 40 AU, respectively.

Initial conditions for the disk's structure and thermodynamic state were set using an equilibrium star plus disk model generated using a modified \citet{hachisu1986} self-consistent field (SCF) relaxation method \citep{pickett1996,pickett2003,mejiaphd2004,mejia2005,cai2006}. Here  the density and angular momentum distributions are iteratively solved, using the specified equation of state,  until convergence is achieved. This  equilibrium disk was then seeded with small 0.01\% $\delta\rho/\rho$ random perturbations to allow GIs to grow.  The resultant marginally unstable disk served as the initial disk for the simulation.

The initially unstable disk model passed through several phases of evolution \citep[see also][]{pickett2003, mejia2005}.  The initial disk is unstable to the growth of non-axisymmetric structures. During the initial axisymmetric phase, 
which lasts several ORPs, the disk cools and contracts, the contraction is small radially but dramatic in the vertical direction. Around 3-4 ORPs the instabilities begin grow to non-linear amplitudes. When this condition is met, the disk undergoes a strong burst of GIs that predominantly manifest themselves in one or two discrete global spiral modes.  The disk expands violently during the burst phase, producing a significant rearrangement of the disk's mass distribution on a timescale of a few ORPs. This is then followed by a period of several ORPs where heating temporarily washes out some of the non-axisymmetry in the disk.  Finally, following this transition phase, the disk settles into a quasi-steady, long-lived asymptotic state of sustained GI-activity over a large part of the disk, with an overall balance of heating and cooling.

Simulations were run with four different angular resolutions, with grid sizes in $(r,\phi,z)$ of  $(j_{max}, k_{max}, l_{max}) =$ (512,64,64), (512,64,128), (512,64,256), and (512,64,512). Because of the computational overhead associated with running each simulation starting from $t = 0$, the $l_{max} =$ 128 simulation was run through the axisymmetric, burst and transition phases. Simulations using $l_{max} =$ 64, 256, and 512 all begin by interpolating the $l_{max} =$ 128 simulation to higher or lower azimuthal resolution at 9.6 ORPs, near the end of the transition
phase. Although this strategy saves a large amount of computational time, it limits the scope of the analysis to the asymptotic phase. By 12 ORPs, all four disks have transitioned to the asymptotic state.  All simulations were 
run through $\sim 18$ ORPs.

Figure \ref{fig:Final_Q} shows the Toomre $Q$, as a function of radius, for all four simulations at $t \sim$ 18 ORP.  Table \ref{tbl:ams} lists $Q_{avg}$, the value of $Q$ of each of the asymptotic disks, time-averaged from 12 -- 19 ORP and spatially averaged over 10 -- 40 AU.  From $\sim$ 12 -- 50 AU, $Q \sim 1.0$ -- 1.4 and thus the disk is subject GIs over this full range.  Interior to $\sim$ 27 AU, the value of $Q$ shows no dependence on
the azimuthal resolution.  From 27 -- 50 AU, lower resolution simulations display lower values of $Q$. {\bf I suppose 27 AU was chosen by eye? Or was there some other analysis. I vote to toss these last two sentences as I'm not convinced w/o a time averaged quantity. Unless this observation is used in further reasoning in the discussion...SM}

>>>>

\usepackage{natbib}
\usepackage{color}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{grffile}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{amssymb}
\usepackage{ulem}
\graphicspath{{figures/}}

\definecolor{magenta}{rgb}{1.0,0.0,0.85}

\newcommand{\ACBc}[1]{{\color{red} #1}}
\newcommand{\TSCc}[1]{{\color{magenta} #1}}
\newcommand{\SMc}[1]{{\color{blue} #1}}

@@

\title{Convergence Studies of Mass Transport in Disks with Gravitational Instabilities.
 \\
I. The Constant Cooling Time Case}
\author{Scott Michael}
\affil{Astronomy Department, Indiana University, Bloomington, IN 47405}
\email{scamicha@astro.indiana.edu}

@@

\author{Aaron C. Boley}
\affil{Department of Astronomy, University of Florida, Gainesville, FL 32611}
\email{aaron.boley@gmail.com}

@@

We conduct a convergence study of a protostellar disk, subject to a constant 
global cooling time and susceptible to gravitational instabilities (GIs), at
a time when heating and cooling are roughly balanced. Our goal is to 
determine the gravitational torques produced by GIs, the level to which transport can
be represented by a simple alpha-disk formulation, and to examine 
fragmentation criteria. Four simulations are conducted, identical except for the number
of azimuthal computational grid points used. A Fourier decomposition of 
nonaxisymmetric density structures in  $\cos(m\phi)$, $\sin(m\phi)$ is performed to evaluate 
the amplitudes $A_m$ of these structures. The $A_m$, gravitational torques, and
the effective \citet{shakura1973} $\alpha$  arising from gravitational stresses
are determined for each resolution. We find nonzero $A_m$ for all $m$-values and
that $A_m$  summed over all $m$ is essentially independent of resolution. Because
the number of measurable $m$-values is limited to half the number of azimuthal
grid points, higher-resolution simulations have a larger fraction of their total
amplitude in higher-order structures. These structures act more locally than
lower-order structures. Therefore, as the resolution increases the total 
gravitational stress decreases as well, leading higher-resolution simulations to experience
weaker average gravitational torques than lower-resolution simulations. The 
effective $\alpha$ also depends upon the magnitude of the stresses, thus
$\alpha_{eff}$ also decreases with increasing resolution. Our converged
$\alpha_{eff}$ is consistent with predictions from
an analytic local theory for thin disks by Gammie (2001), but only over many
dynamic times when averaged over a substantial volume of the disk.

@@

Detailed three-dimensional hydrodynamic modeling of protostellar disks has provided considerable insight into the physical and thermal states of these disks.  However,  computing resources limit these studies to durations spanning timescales much shorter than a disk's evolutionary lifetime. The level of spatial resolution, and hence temporal resolution due to the Courant condition, can play a deterministic role in interpretations of calculations; while lower resolutions allow more extensive spatial and temporal studies, this benefit can come at the sacrifice of important physics.  Convergence studies provide a means 
of deriving important outcomes while assessing the resolution needed to determining these outcomes.

In contrast to sophisticated calculations that include detailed physics, 
approaches exist that replace detailed physics with statistical parameters that encapsulate this physics.  This
allows for calculations covering much longer time frames.  Both detailed and 
parametric
approaches can play important roles in broadening our understanding of complex nonlinear systems, provided a connection between the detailed physics and statistical parameters can be demonstrated and understood. In what follows, we perform convergence 
tests on hydrodynamical disk simulations to examine the applicability of an $\alpha$-disk 
formulation \citep{shakura1973} to  the
evolution of protostellar disks subject to gravitational instabilities.

Gravitational instabilities (GIs) can play an important, and at times dominant, role in driving the structural 
evolution of  protostellar disks \citep[for reviews, see][]{durisen2005, durisen2007}.
Thermal processes play the primary role in regulating the amplitude and outcome of these instabilities \citep{pickett1998,pickett2000,pickett2003,nelson1998,nelson2000a, mejia2005}. A disk's susceptibility to GIs can be parameterized by the Toomre $Q$-parameter \citep{toomre1981};  $Q = c_s \kappa / \pi G \Sigma$, where $c_s$ is the sound speed, $\kappa$ is the epicyclic frequency ($\sim$ the rotation frequency 
$\Omega$ in a nearly Keplerian disk), and $\Sigma$ is the disk surface mass density. For $Q \al$ 1.7,  small density perturbations in a disk grow exponentially on a timescale comparable to the rotation period \citep{durisen2007}. These perturbations manifest themselves as multi-arm spirals with a predominantly trailing pattern that transports angular momentum outward by gravitational torques \citep{larson1984,boss1984,durisen1986,papaloizou1991, laughlin1994,nelson1998,pickett1998}.

@@

%\ACBc{[ACB: I've been thinking about the whole heating-by-pdV-to-stabilize-disks thingy.  I am not so sure it is true! Let's 
%consider a purely isentropic compression.  In this case, $Q= (\gamma K)^{1/2}\Sigma^{\left(2-\gamma\right)/2}\kappa/(\pi G)$. 
%A compression STILL destabilizes the disk.  One needs to change $K$. Have I set up a straw man by restricting myself to 
%thinking of an isentropic compression? I do not think so because pdV should not change entropy. ]}

We note here a caution on semantics.  GIs in disks commonly manifest themselves  as non-axisymmetric density structures whose strengths can be characterized 
by a Fourier decomposition of the density in terms of $\sin m\phi$ and $\cos m\phi$, where 
$m$ is the order of the Fourier term and $\phi$ is the usual 
cylindrical coordinate.  While discrete coherent spiral waves seen in these  disks may well be correctly thought of as eigenmodes, particularly 
at low $m$-values, the analysis necessary to confirm that these structures are, in fact, truly modes is generally lacking.
Power in a specific $m$-value does not imply the existence of that mode, nor does it necessarily represent the strength of a mode that truly exists.  Indeed, a disk with a {\it single} $m = 2$ eigenmode growing to nonlinear amplitudes will exhibit power at all even values of $m$.  A disk with {\it two} nonlinear eigenmodes of
$m = 2$ and 3 will exhibit power at all $m$-values.  
For this reason, in the balance of this paper we will avoid referring to the Fourier terms as modes, but rather will refer to them as Fourier structures or components.   The term {\it mode} will be reserved for those cases where eigenmodes are known, or strongly suspected, to exist.

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime \citep{nelson1998, pickett1998, lodato2004,boley2006,cossins2009}.  Even if the initial growth is dominated by one mode, numerous modes usually appear in the nonlinear regime. As a consequence, if a disk achieves a quasi-steady balance between heating and cooling, referred to as the {\it asymptotic state},
then power will exist at all $m$-values resolved by a given numerical method \citep{mejia2005,boley2006}.  
In the asymptotic state, GI-active disks develop overlapping density structures of different strengths, geometries and coherence that pervade the entire disk \citep[e.g.,][]{pickett2003, mejia2005,boley2006}.
%\citet{gammie2001} refers to this state as {\it gravitoturbulence}. Gravitoturbulence 
These structures provide a possible bridge between the detailed physics of GIs and the viscous transport parameter $\alpha$
\citep{gammie2001}.  Here convergence studies are important because 
higher-order modes dissipate energy on shorter wavelengths than lower-order modes; the mix of low and high-order azimuthal density structures
available to a disk is set by the resolution of the simulation.

\citet{shakura1973} proposed turbulence as the primary source of effective viscosity in gas disks
and parameterized turbulent transport in a steady-state disk with the  parameter $\alpha$.  In this case,  the kinematic
viscosity is defined by $\nu = \alpha c_s H$, where  $c_s$ is the sound speed and $H$ is the disk height.
The source of the 
turbulence in their $\alpha$-disk formulation is unspecified. Although their original work was directed towards X-ray bright
accretion disks around black holes, this heuristic approach has proven fruitful in many astrophysical situations by allowing for analytic solutions and relatively easy numerical modeling over time-spans inaccessible to detailed modeling. 
%In the  $\alpha$-disk formalism, when the accretion rate $\dot{M}$ is constant in both time and radius, 
%$\dot{M} = 3\pi\alpha c_s \Sigma H$ where $\Sigma$ is the surface density and $H$ represents
%the disk scale height. 
The $\alpha$ formalism makes no assumption about the nature of the angular momentum transport, only that {\it angular momentum transport
and heating are dominated by local processes}. \citet{lin1987,lin1990} were among the first to suggest
that GI-induced transport could be described within a viscous
framework.  However, the  validity of an $\alpha$-disk picture requires self-consistency of local energy dissipation. 
The locality of dissipation in GI-active disks has not been rigorously tested and remains an open question.

In fact, several studies support the idea that angular momentum transport by GIs in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.,][]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006}.  
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in rough balance, 
 it may be valid to represent GI-induced transport by an {\it effective viscosity}, as if it were a local process \citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.  In particular, multiple studies suggest that
an effective $\alpha$-viscosity may be consistent
with transport in GI-active disks for a disk to star mass ratio $M_d/M_* \le 0.2$ -- 0.3; GI-active disks
in systems with larger $M_d/M_*$ are dominated by lower-order modes that act in a more global sense and thus cannot be well represented by an $\alpha$-approach
\citep{lodato2004, cossins2009, vorobyov2010}. 
We note, however, that the these studies all assumed
a {\it local cooling time}, $t_{cool}\Omega =$ constant, 
where $t_{cool}$ is defined by the rate at which cooling decreases the disk's internal energy density
$\epsilon$ by $d\epsilon/dt = \epsilon / t_{cool}$.  In contrast, we and our collaborators use either a {\it global cooling time},
$t_{cool} = $ constant \citep[e.g.,][this paper]{pickett2003, mejia2005}, or use realistic opacities \citep[e.g.,][]{mejiaphd2004, cai2006, cai2008, boley2006, boley2007,  boley2009} which lead to something in between $t_{cool} = $ constant  and
$t_{cool}\Omega =$ constant.

\citet{gammie2001} derived an effective alpha for the case of a thin gravitationally unstable disk where the heating caused by GIs was balanced by the local cooling rate.  
Assuming a local cooling time of $t_{cool}\Omega =$ constant,  Gammie found
\begin{equation}
\alpha_{\rm{eff}} = [\gamma(\gamma - 1) \slantfrac{9}{4}\Omega t_{cool}]^{-1},
\label{eq:gammie}
\end{equation}
where $\gamma$ is the two-dimensional adiabatic index. 
The two-dimensional index relates
to the three-dimensional adiabatic index, $\Gamma$, by 
$\gamma = (3\Gamma - 1) / (\Gamma + 1)$
for a non-self-gravitating
disk and
$\gamma = 3 - 2/\Gamma $
for a strongly self gravitating disk.   \cite{lodato2004} derived a similar result when considering three-dimensional protoplanetary disks.
Fully 3D simulations \citep[e.g.,][]{lodato2004, cossins2009} suggest that, at least in the case of imposed 
$t_{cool}\Omega = $ constant cooling, $\alpha_{\rm{\rm{eff}}}$ roughly converges to the value predicted by \citeauthor{gammie2001}.

In what follows, we address these and related questions using a grid-based, finite-difference 3D hydrodynamics code with 
self-gravity to follow the evolution of a protostellar disk with a constant cooling time and subject to GIs.
We examine how grid resolution affects
the amplitude distribution of azimuthal density structures arising from GIs, the gravitational torques arising from these structures,
the effective $\alpha$, and the extent to which transport can be represented by an $\alpha$-disk formulation.
%For computational grids of a fixed {\it physical} size,
%grids with more elements provide finer meshes, and thus a more accurate solution. 
%However, increased accuracy is offset by computational penalties.
%The number of equations solved at each time step 
%increases as the grid size is increased.  At the same time, the Courant condition 
%forces shorter time steps.  
A mesh convergence study 
provides a common approach to balancing accuracy and computing resources.
It permits an assessment of whether or not convergence occurs and what resolution is required.
%allowing us to determine if convergence occurs and what is the required resol
In the following sections, we follow the evolution of a disk subject to GIs
using grids with four different
resolutions in the azimuthal direction.  The number of azimuthal elements used in simulations
may be particularly important, because higher azimuthal resolution 
allows power in non-axisymmetric density structures to spread to higher-order modes 
that behave more locally. 
Although full convergence testing eventually requires investigation of the radial and vertical resolution in our
cylindrical grid, these tests are expensive and will be left to a future paper.  We suspect that the azimuthal
direction is most important to test first because it is non-axisymmetry that produces the torques leading
to mass and angular momentum transport.
We test this by examining the amplitude in various 
$m$-values to see how they are affected by the choice of grid.
In the process, 
we report the results of a convergence study designed to determine whether GI-active disks, chaotic by nature, achieve statistical equilibrium.   If this is the case, we investigate how closely these disks approximate equation (1) from
 \citet{gammie2001} and examine the distribution of $\alpha_{\rm{eff}}$.

The balance of this paper is organized as follows.  Section 2 provides the details of the numerical approach and defines the models.  An analysis of the azimuthal density structures of the simulations 
is found in \S3.1, the gravitational torques arising from these structures follows in \S3.2.  Effective converged alphas are 
discussed in \S3.3.  Section 3.4 deals with the degree to which selected grid-sizes lead to results consistent with
convergence.  Fragmentation is discussed in \S3.5.  A summary of results is found in \S4.

@@

Simulations were conducted with the three-dimensional hydrodynamics code used by the Indiana University Hydrodynamics Group in several previous studies \citep{pickett1998, pickett2000, mejiaphd2004, mejia2005, boley2006, cai2008}.  This code uses a second-order, explicit, Eulerian scheme to solve Poisson's equation, an ideal gas equation of state, and the equations of hydrodynamics in conservative form on a uniform cylindrical grid.  The code assumes mirror symmetry about the equatorial plane. It includes both self-gravity and artificial viscosity; the latter serves to 
mediate shocks. Source and flux terms \citep{norman1986} are computed separately in an explicit, second-order time integration \citep{albada1982,christodoulou1991,yangphd1992}.

Initial conditions for the disk's structure and thermodynamic state were set using an equilibrium star plus disk model generated by a modified \citet{hachisu1986} self-consistent field relaxation method \citep{pickett1996,pickett2003,mejiaphd2004,mejia2005,cai2006}. Here  the density and angular momentum distributions are iteratively solved, using the specified equation of state,  until convergence is achieved. This  equilibrium disk was 
given random cell-to-cell $\sim10^{-4}$ density perturbations to seed growth of  GIs.  The resultant marginally unstable disk served as the initial disk for the simulation.

Because this is a polytropic disk with an idealized cooling time, only the ratio of disk to stellar mass, given here by 0.153, is specified.  In what follows, quantities specified with real physical units are derived assuming a 1 $M_\odot$ central star and a 0.153  $M_\odot$ disk. In this case, the outer rotation period (ORP) is defined at $r = 33$ AU and one ORP $\approx$ 214 years.  The disk has an initial surface density that approximates $\Sigma \sim r^{ -1}$, with initial inner and outer radii of 2.3 and 40 AU, respectively.

 We assume an equation of state $P = (\gamma -1)\epsilon$, where $P$ is the pressure, $\epsilon$ is the internal energy density, and $\gamma$ is the ratio of specific heats, given here by $\gamma = 5/3$.  The model disks are cooled by decreasing their internal energy according to the prescription $d\epsilon/dt = \epsilon / t_{cool}$, where $t_{cool}$ is the global cooling time, which is set to 2 ORP everywhere.

The initially unstable disk model passed through several phases of evolution \citep[see also][]{pickett2003, mejia2005}.  
During the initial axisymmetric phase, 
which lasts several ORPs, the disk cools and contracts.  The contraction is small radially but dramatic in the vertical direction. Around 3 to 4 ORPs, the instabilities begin to grow to nonlinear amplitudes, and the disk 
then undergoes a strong burst of GIs that predominantly manifest themselves in a few discrete global spiral modes.  The disk expands violently during the burst phase, producing a significant rearrangement of the disk's mass distribution on a timescale of a few ORPs. This is then followed by a period of several ORPs where heating temporarily washes out some of the non-axisymmetry in the disk.  Finally, following this transition phase, the disk settles into a quasi-steady, long-lived asymptotic 
phase of sustained GI-activity over a large part of the disk, with an overall balance of heating and cooling.

Simulations were run with four different angular resolutions, with grid sizes in cylindrical coordinates $(r,\phi,z)$ of  $(j_{max}, k_{max}, l_{max}) =$ (512,64,64), (512,64,128), (512,64,256), and (512,64,512). 
For reference, the radius $r = 33$ AU at which the ORP is defined is at radial cell $j = 200$.  The radial grid extends to 512
to accomodate disk expansion during outburst.
Because of the computational overhead associated with running each simulation starting from $t = 0$, the $l_{max} =$ 128 simulation was run through the axisymmetric, burst and transition phases. Simulations using $l_{max} =$ 64, 256, and 512 all begin by a linear interpolation of the $l_{max} =$ 128 simulation to higher or lower azimuthal resolution at 9.6 ORPs, near the end of the transition
phase. Although this strategy saves a large amount of computational time, it limits the scope of the analysis to the asymptotic phase. By 12 ORPs, all four disks appear to have transitioned to the asymptotic state,
e.g., they all display roughly steady $Q(r)$ profiles starting at $\sim 12$ ORPs.  All simulations were 
run through $\sim 18$ ORPs.

Figure \ref{fig:Final_Q} shows the Toomre $Q$, as a function of radius, for all four simulations at $t \sim$ 18 ORP.  
%Table \ref{tbl:ams} 
Table 1 lists $Q_{avg}$, the arithmetic mean of $Q$ for each of the asymptotic disks, time-averaged from 12 -- 18 ORP and spatially averaged over 10 -- 40 AU.  From $\sim$ 12 -- 50 AU, $Q \sim 1.0$ -- 1.4 and thus the disk is subject to GIs over this full range.
